\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-07 ace-pldi14/ace-pldi14.tex
Finished abstract of PLDI14

@@

\maketitle
\begin{abstract}
Evidence suggests that programmers are unwilling to adopt new languages to gain access to language-integrated abstractions, even when they agree that these abstractions might be valuable if all else were held equal. This suggests that language designers should give greater consideration to two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems, and \emph{extensibility}, so that adopting a new language-integrated abstraction does not require adopting a new language.

In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for the world's most widely-used dynamically-typed language, Python. 
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}

>>>>

\usepackage{todonotes}

@@

\maketitle
\begin{abstract}
Evidence suggests that programmers are reluctant to adopt new programming languages to gain access to new abstractions, even when they agree that these abstractions could be valuable for their work. This suggests a need for languages designed to satisfy two related design criteria: \emph{compatibility} with existing languages, tools and ecosystems; and \emph{extensibility}, so that adopting new abstractions does not require adopting a new language.

In this paper, we introduce \Ace, a language compatible with the tools and infrastructure developed for Python, among the most  widely-adopted dynamically-typed language today. While Python and other similar languages are considered satisfactory for small-scale scripting tasks, \Ace~is designed for situations where correctness and performance are particularly important, offering a  static type system and more granular control over compilation. Unlike most statically-typed languages, however, the \Ace~type system can be extended from within. The extension mechanism, called \emph{active typechecking and translation}, guarantees that extensions cannot interfere  with one another by restricting the scope of extension logic, which must be associated with a type, around expressions of that type. We show that this mechanism remains powerful enough to express the static and dynamic semantics of several common functional, object-oriented and parallel programming primitives, as well as varied domain-specific abstractions, all from within libraries. 
Ace can be used from the shell or interactively from within Python\todo{take this sentence out?}.
%Researchers developing languages and abstractions for high-performance computing must consider a number of design criteria, including performance, verifiability, portability and ease-of-use. Despite the deficiencies of legacy tools and the availability of seemingly superior options, end-users have been reluctant to adopt new language-based abstractions. We argue that this can be largely attributed to a failure to consider three additional criteria: continuity, extensibility\- and interoperability. This paper introduces Ace, a language that aims to satisfy this more comprehensive set of design criteria. To do so, Ace introduces several novel compile-time mechanisms, makes principled design choices, and builds upon existing standards in HPC, particularly Python and OpenCL. OpenCL support, rather than being built into the language, is implemented atop an extensibility mechanism that also admits abstractions drawn from other seemingly disparate paradigms. The core innovation underlying this and other features of Ace is a novel reification of types as first-class objects at compile-time, representing a refinement to the concept of active libraries that we call \emph{active types}. We validate our overall design by considering a case study of a simulation framework enabling the modular specification and efficient execution of ensembles of neural simulations across a cluster of GPUs.
\end{abstract}
